---
title: "gmtol tree"
output: html_document
date: "2024-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
install.packages("ape")
library(ape)
library(phangorn)
install.packages("phytools")
library(phytools)
install.packages('taxize')
library(taxize)
```
```{r github-installation, eval=FALSE} 
library(devtools)
install_github("arendsee/phylostratr")
library(phylostratr)


if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("ggtree")

library(ggtree)
library(ape)
```

```{r}
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL")
gl=readLines("GMTOL tree listAug8taxid.txt")

ncbi_tree(gl)


```
doesnt seem to work well
now trying chatgpt R method

```{r}
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL")
taxa_ids <- readLines("GMTOL tree listAug8taxid.txt")
```

```{r}
taxa_info <- classification(taxa_ids, db = "ncbi")
```

```{r}
tree <- class2tree(taxa_info)
```
it works now! have to do tree$phylo
```{r}
class(tree)
plot(tree)
tree$phylo
dist_matrix <- taxdist(taxa_info)
treephy <- ape::as.phylo(tree)

ggtree(tree$phylo)
tp=tree$phylo
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL")
ape::write.tree(tree$phylo, file = "GMTOLall.newick")
```

```{r}
ggtree(tp, layout="circular")

```
```{r}
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL/phylo")
library(ape)
library(phytools)
install.packages('castor')
library(castor)

```
```{r}
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL/phylo")
table=read.table("GrpTablePhylum2.txt",sep='\t',header=TRUE,row.names=1)
htree=read.tree("hostf_tree2.nwk")


```

destroyX = function(es) {
  f = es
  for (col in c(1:ncol(f))){ #for each column in dataframe
    if (startsWith(colnames(f)[col], "X") == TRUE)  { #if starts with 'X' ..
      colnames(f)[col] <- substr(colnames(f)[col], 2, 100) #get rid of it
    }
  }
  assign(deparse(substitute(es)), f, inherits = TRUE) #assign corrected data to original name
}

destroyX(table)
head(table)

```{r}
length(htree$tip.label)

nrow(table)

head(htree$tip.label)

rownames(tablet) <- sub(".","_",rownames(tablet))  #do all this if needed. I edned up importing and fixing in excel

htips=htree$tip.label
irow=rownames(table)

int=intersect(irow,htips)
length(int)
table2=table[int,]
nrow(table2)

phylopca<-phyl.pca(htree,table2)
print(phylopca)
```

```{r}
biplot(phylopca,cex=0.3)
```
```{r}
install.packages('factoextra')
install.packages('ggbiplot')
library(ggbiplot)
library(factoextra)
```

```{r}
install.packages('plotly')
install.packages('ggfortify')
library(plotly)
library(ggfortify)
```


```{r}
prin=as.princomp(phylopca)
prc=as.prcomp(phylopca)
prin$scale=prin$scale/10
prin$scores
```

trying to merge metadata with prin

```{r}
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL/phylo")
meta=read.table("GrpSpeciesMetadataAug21_hostf.txt",sep='\t',header=TRUE, row.names=1)
```
```{r}
length(prin$scores)
# Load necessary libraries
library(factoextra)

# Assuming 'prin' is your PCA object and 'metadata' is your metadata dataframe

# Merge PCA results with metadata
int=intersect(rownames(meta),rownames(prin$scores)
meta2=meta[int,]
merged_data <- merge(as.data.frame(prin$scores), meta, by = "row.names", all.x = TRUE)
rownames(merged_data) <- merged_data$Row.names
merged_data$Row.names <- NULL

# Now plot the PCA biplot with colored points based on a metadata column (e.g., 'group')

#this is to add a Diet column to PCA data..this can work for so many things. Vlookup basically

p$data
p$data$DietSimp = meta2$DietSimp[match(rownames(p$data),rownames(meta2))]
```
```{r}
p=(fviz_pca_biplot(prin, 
                select.var = list(contrib = 5),
                geom = c("point", "text"), 
                      geom.var = c("point", "text"),
                      palette="ucscgb",
                      label="var",       
                arrowsize = 1, 
                labelsize = 0.5,
                 repel = TRUE,
                invisible="quali",
                      pointshape = 19, pointsize = 1.5, alpha=0.7,
                col.ind = meta2$Class,xlim=c(-1100,700),ylim=c(-1000,1000)))
 labs(colour= "Class" )
```

```{r}
ggplot(p$data,
aes(x=x,y=y,color=Col.,shape=DietSimp)) + geom_point()+
                      theme_minimal() +
                     theme(plot.title = element_text(hjust = 0.5,size = 20, face = "bold"),
                           legend.title = element_text(size = 12), 
                           legend.text = element_text(size = 8),
                           legend.key.size = unit(0.5,"line") )+
                     guides(shape = guide_legend(override.aes = list(size = 4)),
                           color = guide_legend(override.aes = list(size = 4)))+
                             scale_shape_manual(values=c(16,16,16,17,16))
```


p$layers[[1]]$meta2$DietSimp <- factor(meta2$DietSimp)
p$layers[[1]]$mapping <- aes(x, y, colour = meta2$Class, shape = meta2$DietSimp)
p$layers[[1]]$aes_params$size <- 1
p$layers[[1]]$aes_params

p <- p + labs(shape = 'DietSimp')
p
```


```{r}
p=fviz_pca_biplot(prin, select.var = list(contrib = 5),geom=("point"),geom.var=c("arrow","text")
      , arrowsize = 1, labelsize=0.5   ) 
p+geom_segment(data = prin$loadings, aes(x = 0, y = 0, xend = (Comp.1*5),
     yend = (Comp.2*5)), arrow = arrow(length=prin$loadings))+
  annotate("text", x = (prin$loadings$Comp.1*5), y = (prin$loadings$Comp.2*5),
     label = PCAloadings$Variables)
```

```{r}
fviz(prc, "ind")
p=fviz_pca_biplot(prc, select.var = list(contrib = 5))
p+geom_segmen(arrow = arrow(length = unit(1/2)))
```
install.packages('PHclust')
```{r}
library(PHclust)
 Hybrid(table, absolute = TRUE,Kstart=4,Treatment = rep(c(1,2), each = 2))

```

Trying to calc tree distance from humans

```{r}
library(ape)
library(phytools)
library(ggtree)
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL/phylo")
tt=read.tree("hostf_tree2.nwk")

ggtree(tt, layout="circular")
cot=cophenetic(tt)
cot
View(cot)

which(colnames(cot) == "Homo_sapiens")
235
coth=cot[,235,drop=FALSE]
head(coth)
View(coth)
length(coth)

tabler=table

#library(phyloseq)
#tabler = transform_sample_counts(table, function(x) x / sum(x) ) # doesnt work for some reason
head(table)
tabler <-sweep(tabler, 2, colSums(tabler), `/`)
head(tabler)
tail(table)

which(colnames(tabler) == "d__Bacteria.p__Firmicutes_A")
2
tableFirmr=tabler[,2,drop=FALSE]

cothtabler=transform(merge(coth,tableFirmr,by=0,all=TRUE), row.names=Row.names, Row.names=NULL)
head(cothtabler)

plot((cothtable[,1]),(cothtable[,2]/1000),xlab="Distance to humans",ylab="Rel. Abund. of Firmicutes Clade A")

model <- lm((cothtabler[,2]) ~ (cothtabler[,1]), data = cothtabler)
summary(model)
```



```{r}
d__Bacteria.p__Firmicutes_D is column 5
cothtable5=transform(merge(coth,table[,1:5],by=0,all=TRUE), row.names=Row.names, Row.names=NULL)
head(cothtable5)

p1=plot((cothtable5[,1]),(cothtable5[,3]/1000),xlab="Distance to humans",ylab="Rel. Abund. of Firmicutes Clade A")
p2=plot((cothtable5[,1]),(cothtable5[,6]/1000),xlab="Distance to humans",ylab="Rel. Abund. of Firmicutes Clade D")
#Clade D doesnt have much of a trend
p3=plot((cothtable5[,1]),(cothtable5[,2]/1000),xlab="Distance to humans",ylab="Rel. Abund. of Proteobacteria",col=(alpha))

par(mfrow = c(1, 3))




```
wana just see distributuion of class counts to see if i have good primitive coverage
```{r}
meta2
tapply(rownames(meta2),meta2$Class, FUN = function(x) length(unique(x)))
```
fyi using UCSD now but some stuff is still on northwestern
setwd("/Users/samd/Library/CloudStorage/OneDrive-UniversityofCalifornia,SanDiegoHealth/TOL2024/phylo")
```{r}
library(dplyr)
meta2=as.data.frame(meta2)
meta2$sampleid=rownames(meta2)
cothtable5$sampleid=rownames(cothtable5)
```

```{r}
replicate(999,
  {
      m21=meta2 %>% 
        group_by(Class) %>% 
        slice_sample(n=21) |>
        ungroup()## or we can use: filter(row_number() <= 5)
      m21
   cothtable5f = merge(cothtable5, m21, by='sampleid')
  
   model <- lm(d__Bacteria.p__Firmicutes_A ~ Homo_sapiens, data = cothtable5f)
summary(model)


r2=summary(model)$r.squared
cb=cbind(coef(summary(model)),r2)

 write.table(cb, 'FirmA_vsHom.csv' , append = TRUE, sep = ",", col.names = FALSE,row.names=TRUE) 
  }
)
```


```{r}
#geom code for Firmicutes

gF=ggplot(cothtable5f, aes(x=Homo_sapiens, y=d__Bacteria.p__Firmicutes_A/1000)) +
  geom_point(alpha=0.3) +
  ylim(0,1.0)+
  ylab("Rel. Abund. of Firmicutes")+
  xlab('Divergence from Homo sapiens')+
   theme_classic()+
  geom_smooth(method=lm , color="gray2", fill="gray2", se=TRUE)+
    theme(axis.title.y = element_text(color="black",size=12))+
     theme(axis.title.x = element_text(color="black",size=12))+
     theme(axis.text.x = element_text(color="black",size=10),
         axis.text.y = element_text(color="black", 
                                    size=10))+
                                    theme(plot.margin = unit(c(0.8,0.8,0.8,0.8), "cm"))
                                    
 
  gF
```

```{r}
#Proteobacteria
```


```{r}
gP=ggplot(cothtable5f, aes(x=Homo_sapiens, y=d__Bacteria.p__Proteobacteria/1000)) +
  geom_point(alpha=0.5) +
  ylab("Rel. Abund. of Proteobacteria")+
  xlab('Divergence distance from Homo sapiens')+
  geom_smooth(method=lm , color="gray2", fill="gray2", se=TRUE)+
  theme_classic()
  gP
```

```{r}
gB=ggplot(cothtable5f, aes(x=Homo_sapiens, y=d__Bacteria.p__Bacteroidota /1000)) +
  geom_point(alpha=0.5) +
  ylab("Rel. Abund. of Bacteroidota")+
  xlab('Divergence distance from Homo sapiens')+
  geom_smooth(method=lm , color="gray2", fill="gray2", se=TRUE)+
  theme_classic()
  gB
```


```{r}
replicate(999,
  {
      m21=meta2 %>% 
        group_by(Class) %>% 
        slice_sample(n=21) |>
        ungroup()## or we can use: filter(row_number() <= 5)
      m21
   cothtable5f = merge(cothtable5, m21, by='sampleid')
  
   modelp <- lm(d__Bacteria.p__Proteobacteria ~ Homo_sapiens, data = cothtable5f)
summary(modelp)


r2=summary(modelp)$r.squared
cbp=cbind(coef(summary(modelp)),r2)

 write.table(cbp, 'FirmA_vsProt.csv' , append = TRUE, sep = ",", col.names = FALSE,row.names=TRUE) 
  }
)
```

```{r}
  modelp <- lm(d__Bacteria.p__Proteobacteria ~ Homo_sapiens, data = cothtable5f)
summary(modelp)
```
```{r}
using UCSD now
setwd("/Users/samd/Library/CloudStorage/OneDrive-UniversityofCalifornia,SanDiegoHealth/TOL2024/phylo")
```
now i wanna add class data to this and add Class plots

```{r}
setwd("/Users/samd/Library/CloudStorage/OneDrive-UniversityofCalifornia,SanDiegoHealth/TOL2024/phylo")
tableClass=read.table("GMTOLsong_table2024_N20_f2all_V4_GrpSpecies_Class.tsv",sep='\t',header=TRUE,row.names=1)
tableClass=destroyX(tableClass)
tableClass
tclass=as.data.frame(t(tableClass))
head(tclass)
View(tclass)
rownames(tclass) <- gsub("\\.", "_", rownames(tclass))


```

```{r}
which(colnames(tclass) == "d__Bacteria;p__Firmicutes_A;c__Clostridia_258483")
3
tclassClost=tclass[,3,drop=FALSE]
tclassBact=tclass[,2,drop=FALSE]
tclassAlph=tclass[,4,drop=FALSE]
#adding Clost to cothtable
cothtableClost=transform(merge(cothtabler,tclassClost,by=0,all=TRUE), row.names=Row.names, Row.names=NULL)
cothtableBact=transform(merge(cothtableClost,tclassBact,by=0,all=TRUE), row.names=Row.names, Row.names=NULL)
cothtableAlph=transform(merge(cothtableBact,tclassAlph,by=0,all=TRUE), row.names=Row.names, Row.names=NULL)
head(cothtableClost)
View(cothtableAlph)

cothtableAlph$sampleid=rownames(cothtableAlph)

```
names(cothtableAlphf) <- gsub("\\.", "_", names(cothtableAlphf))
```{r}
m21a=meta2 %>% 
        group_by(Class) %>% 
        slice_sample(n=21) |>
        ungroup()## or we can use: filter(row_number() <= 5)
      m21a
   cothtableAlphf = merge(cothtableAlph, m21a, by='sampleid')
  
   modelClost <- lm(d__Bacteria.p__Firmicutes_A_c__Clostridia_258483 ~ Homo_sapiens, data = cothtableAlphf)
summary(modelClost)

gClost=ggplot(cothtableAlphf, aes(x=Homo_sapiens, y=d__Bacteria.p__Firmicutes_A.c__Clostridia_258483/1000)) +
  geom_point(alpha=0.5) +
  ylab("Rel. Abund. of Clostridia")+
  xlab('Divergence distance from Homo sapiens')+
  geom_smooth(method=lm , color="gray2", fill="gray2", se=TRUE)+
  theme_classic()
  gClost

gClost=ggplot(cothtableAlphf, aes(x=Homo_sapiens, y=d__Bacteria.p__Firmicutes_A.c__Clostridia_258483/1000)) +
  geom_point(alpha=0.5) +
  ylab("Rel. Abund. of Clostridia")+
  xlab('Divergence distance from Homo sapiens')+
  geom_smooth(method=lm , color="gray2", fill="gray2", se=TRUE)+
  theme_classic()
  gClost



gBact=ggplot(cothtableAlphf, aes(x=Homo_sapiens, y=d__Bacteria.p__Bacteroidota.c__Bacteroidia/1000)) +
  geom_point(alpha=0.5) +
  ylab("Rel. Abund. of Bacteroidia")+
  xlab('Divergence distance from Homo sapiens')+
  geom_smooth(method=lm , color="gray2", fill="gray2", se=TRUE)+
  theme_classic()
  gBact
  
   
gAlph=ggplot(cothtableAlphf, aes(x=Homo_sapiens, y=d__Bacteria.p__Proteobacteria.c__Alphaproteobacteria/1000)) +
  geom_point(alpha=0.5) +
  ylab("Rel. Abund. of Alphaproteobacteria")+
  xlab('Divergence distance from Homo sapiens')+
  geom_smooth(method=lm , color="gray2", fill="gray2", se=TRUE)+
  theme_classic()
  gAlph


plot_grid(gP,gF,gB,gAlph,gClost,gBact)
  



plot((cothtableAlphf[,2]),(cothtableAlphf[,4]/1000),xlab="Distance to humans",ylab="Rel. Abund. of Clostridia")

model <- lm((cothtabler[,2]) ~ (cothtabler[,1]), data = cothtabler)
summary(model)
```

Checking how many species I have per primer

```{r}
setwd("/Users/samd/Library/CloudStorage/OneDrive-UniversityofCalifornia,SanDiegoHealth/TOL2024/phylo")
tm=read.table("GrpSpeciesMetadataDec4.txt",sep='\t',header=TRUE)
tm
tapply(tm$prefname,tm$Primer2, FUN = function(x) length(unique(x)))
```



Now trying to see what this looks like with insects as basal

```{r}
library(ape)
library(phytools)
library(ggtree)
setwd("/Users/samd/OneDrive\ -\ Northwestern\ University/TOL/phylo")
tt=read.tree("hostf_tree2.nwk")

ggtree(tt, layout="circular")
cot=cophenetic(tt)
head(cot)
View(cot)
dim(cot)
Temnothorax_nylanderi  #trying ants because other stuff like worms are so isolated evolutionary that everything will be distant

which(colnames(cot) == "Temnothorax_nylanderi")
View(cot[,565])
565
coto=cot[,565,drop=FALSE]
head(coto)

length(coth)


#seems  like this is waste and will do tree like ROb suggested

d__Bacteria.p__Firmicutes_D  d__Bacteria.p__Proteobacteria  d__Bacteria.p__Bacteroidota d__Bacteria.p__Firmicutes_A

tablephy=table2[,c(1,2,3,5),drop=FALSE]
head(tablephy)
write.csv(tablephy,"tablephy_barchart.csv")

```



```{r}
install.packages("doParallel")
devtools::install_github("wdl2459/ConQuR")

```

```{r}
setwd("/Users/samd/Library/CloudStorage/OneDrive-UniversityofCalifornia,SanDiegoHealth/TOL2024/batcheffects")
```

now trying with new table based on v3v4 and v4 and new tree with timetree names resolved

```{r}
setwd("/Users/samd/Library/CloudStorage/OneDrive-UniversityofCalifornia,SanDiegoHealth/TOL2024/phylo")
pgrp=as.data.frame(read.table("GrpTimeTree2_Phylum_GMTOL_unfilt.txt",sep='\t',header=TRUE))
htre=read.tree("testtreeFeb17.nwk")
plot(htre)
pgrp$sampleid <- sub(" ", "_", pgrp$sampleid)
rownames(pgrp)=pgrp$sampleid
pgrp=pgrp[,-1]
head(pgrp)

```

destroyX = function(es) {
  f = es
  for (col in c(1:ncol(f))){ #for each column in dataframe
    if (startsWith(colnames(f)[col], "X") == TRUE)  { #if starts with 'X' ..
      colnames(f)[col] <- substr(colnames(f)[col], 2, 100) #get rid of it
    }
  }
  assign(deparse(substitute(es)), f, inherits = TRUE) #assign corrected data to original name
}

destroyX(table)
head(table)

```{r}
length(htre$tip.label)

nrow(pgrp)

head(htre$tip.label)

htips=htre$tip.label
irow=rownames(pgrp)

int=intersect(irow,htips)
length(int)
table2=pgrp[int,]
nrow(table2)

#prune tree to match table! nice code

htre2<-drop.tip(htre,htre$tip.label[-match(int, htre$tip.label)])
length(htre2$tip.label)
plot(htre2)
ape::write.tree(htre2, file = "GMTOLallFeb17.newick")
```

now making tree without V1-V3 primers
```{r}
setwd("/Users/samd/Library/CloudStorage/OneDrive-UniversityofCalifornia,SanDiegoHealth/TOL2024/phylo")
gmt=read_tsv("GrpSpeciesMetadataFeb20_25.txt")
gmt

# using subset function
gmt2 <- subset(gmt, Primer2=="V1-V3" & reads!="NA" & Genus=="Cephalotes")
gmt2
print(gmt2,n=50)
antlist=(gmt2[,"timetree3"])
antlist2=print((antlist))

htre3<-drop.tip(htre2,htre2$tip.label[match(antlist$timetree3, htre2$tip.label)])
length(htre3$tip.label)
plot(htre3, "f")
ape::write.tree(htre3, file = "GMTOLallFeb20_no_ants.newick")
```

